What We Eval in the Shadows

### 2021-04-17 paper/main.tex
5.4 wip

The code executed by \eval can do side effects. From the compiler perspective,
we care about the observable side effects in environments---\ie variable
definitions, updates, and removals that are visible after a call to \eval
finishes. Knowing where---\ie in which environments do these side-effects
happen, can help us determine how much of the compiler knowledge about the
program will be potentially invalidated.

The \rdyntrace contains low-level hooks capturing the life-cycle of an
environment and variables defined in it. We register callbacks in the tracer
that records each environment creation and removal with updates to its
bindings. From the recorded data, we ignore side effects coming from \eval
sites from unit testing frameworks.\footnote{In the corpus, we have \c{RUnit,
testthat, tinytest} and \c{unitizer} unit testing frameworks.} They run the
testing code via \eval, and thus the results would be biased because of the
high number of code in these tests (\CranRunnableCodeTestsRnd lines).

From the \packageNbrunsRnd programs, we capture \SEAllRnd side effects from
\SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
(\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
side effects caused by the R virtual machine implementation that are not
related to the user code. For example, the \c{.Random.seed} variable, which
contains the state of the random number generator, is saved and restored from
and to the global environment every time the user calls one of the base
routines that use random numbers.
%
% \begin{wrapfigure}{r}{5cm} \hspace*{-20mm} \centering
% \includegraphics[width=90mm]{se-types.pdf} \caption{Distribution of R types
% in target environments for side effects.} \label{fig:se-types}
% \end{wrapfigure}
%
Removing them leaves us with \SEUserRnd side effects (\SEUserRatio) from
\SEUserCallsRnd \eval calls (\SEUserCallsRatio) in \SEUserSites sites
(\SEUserSitesRatio). These sites are part of \SEUserFunctions R functions
(\SEUserFunctionsToAllRatio of all functions doing \eval) in \SEUserPackages
packages. From these functions, \SEFunsNighty are responsible for 90\% of side
effects. Half of the side-effects come just from three functions:
\c{plyr::allocate\_column} (allocates space for a new data frame column),
\c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
effects in the environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally, \SESitesBothEnvirRatio does both.

>>>>

\Eval may peform side-effects. From a compiler's perspective, we care about
effects that can be observed --- \ie variable definitions, updates, and removals
that are visible after a call finishes. Knowing in which environments these
side-effects happen, can help us determine how much of the compiler knowledge
about the program will be invalidated. Our analysis records information about
every environment. From the recorded data, we discard side-effects coming from
unit testing frameworks.\footnote{In the corpus, we have \c{RUnit, testthat,
  tinytest} and \c{unitizer} testing frameworks.} They run their tests via
\eval, and thus everything becomes a side-effect.

From our 98K
%\packageNbrunsRnd
programs, we capture
14M
%\SEAllRnd
side-effects from
\SEAllCallsRnd \eval calls (\SEUserCallsToAllRatio) in \SEAllSites sites
(\SEUserSitesToAllRatio). Again, the challenge is to remove the accidental
side-effects caused by the R virtual machine implementation which are not
related to the user code. For example, the \c{.Random.seed} variable, which
contains the state of the random number generator, is saved and restored from
and to the global environment every time  a statistical routine is called.
%
Removing those leaves us with \SEUserRnd side effects from \SEUserCallsRnd \eval
calls in \SEUserSites sites. These sites are part of \SEUserFunctions R
functions (\SEUserFunctionsToAllRatio of all functions doing \eval) in
\SEUserPackages packages. From these functions, \SEFunsNighty are responsible
for 90\% of side effects. Half of the side-effects come just from three
functions: \c{plyr::allocate\_column} (allocates space for a new data frame
column), \c{withr::execute\_handlers} (executes deferred expressions), and
\c{foreach::doSEQ} (executes an expression on each element in a collection,
possibly in parallel). Most of the \eval sites (\SESitesInEnvirRatio) do side
effects in the environment specified by the \c{envir} parameter (\cf
Section~\ref{sec:eval-in-r}), \SESitesNotInEnvirRatio modifies other
environments, and finally, \SESitesBothEnvirRatio does both.
